Experience Report: Haskell in Computational Biology

### 2012-07-02 mrfy_experience_report.tex
page 1, column 2

Proteins %%% are cellular machinery. They
interact with one another and with other 
molecules to carry out the functions of living cells: metabolism, regulation, 
signaling, and so on.
A~protein's function is determined by its structure, 
and its structure is determined by the sequence of amino acids that
form the protein.
The amino-acid sequence is ultimately determined by a sequence of
nucleic acids in DNA, which we call a gene.
Given a genetic sequence, biologists wish to know the cellular
function of the protein that gene codes for.
One of the best known methods of discovering such function is
to find other proteins of 
similar structure, which likely share similar function.
Proteins that share structure and function are expected to be
descended from a common ancestor---in biological terms, \emph{homologous}---and
thus
the problem of identifying proteins similar to a \textit{query sequence} is called 
\textit{homology detection}.

Computational biologists detect homologies by building 
algorithms which, given a {query sequence}, %% of amino acids,
find known proteins of similar structure.
When similar proteins are formed from amino-acid sequences that
are not too different from the query sequence, the homology can be
detected by
a family of algorithms called 
\textit{hidden Markov models}~\cite{Eddy:1998ut}.
But in real biological systems,
proteins with similar structure and function may be formed from significantly 
different amino-acid sequences, which are not close in edit distance.
Our~research software, MRFy (pronounced
``Murphy''), can detect homologies 
in amino-acid sequences that are only distantly related.
MRFy is available at \url{mrfy.cs.tufts.edu}.
%
%We will explain an 
%algorithm for detecting reasonably similar sequences, and then move on to 
%explain MRFy, our novel approach for detecting more distantly related sequences 
%for proteins that share similar structure and function.
%

>>>>

Proteins, by
interacting with one another and with other 
molecules, carry out the functions of living cells: metabolism, regulation, 
signaling, and so on.
A~protein's function is determined by its structure, 
and its structure is determined by the sequence of amino acids that
form the protein.
The amino-acid sequence is ultimately determined by a sequence of
nucleic acids in DNA, which we call a gene.
Given a gene, biologists wish to know the cellular
function of the protein the gene codes for.
One of the best known methods of discovering such function is
to find other proteins of 
similar structure, which likely share similar function.
Proteins that share structure and function are expected to be
descended from a common ancestor---in biological terms, \emph{homologous}---and
thus
the problem of identifying proteins similar to a \textit{query sequence} is called 
\textit{homology detection}.

Computational biologists detect homologies by building 
algorithms which, given a {query sequence}, %% of amino acids,
compare it with known proteins.
When the known proteins have amino-acid sequences that
are not too different from the query sequence, homology can be
detected by
a family of algorithms called 
\textit{hidden Markov models}~\cite{Eddy:1998ut}.
But in real biological systems,
proteins with similar structure and function may be formed from significantly 
different amino-acid sequences, which are not close in edit distance.
Our~research software, MRFy (pronounced
``Murphy''), can detect homologies 
in amino-acid sequences that are only distantly related.
MRFy is available at \url{mrfy.cs.tufts.edu}.
%
%We will explain an 
%algorithm for detecting reasonably similar sequences, and then move on to 
%explain MRFy, our novel approach for detecting more distantly related sequences 
%for proteins that share similar structure and function.
%
